---
title: "CD2H NLP Sandbox: Format 2014 i2b2 De-id Challenge Data"
author: "Thomas Schaffter, thomas.schaffter@sagebionetworks.org"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output: html_notebook
---

```{r}
rm(list=ls())

# library(dplyr)
# library(purrr)
# library(tibble)
# library(tidyr)
# library(xml2)
```

# Overview

This notebook read the data from the [2014 i2b2 de-identification challenge](https://portal.dbmi.hms.harvard.edu/projects/n2c2-2014/) and export them to the NLP Sandbox De-id format.

# Data

## Input data

Here are the data that we use:

- Test Data: testing-PHI-noTags (testing-PHI-noTags.tar.gz). Each XML file includes the clinical note in XML format.
- Test Data: testing-PHI-Gold-fixed (testing-PHI-Gold-fixed.tar.gz). Each XML file includes the clinical note and the reference annotations.

```{r}
i2b2_input_dir <- file.path("data", "testing-PHI-noTags")
i2b2_gold_dir <- file.path("data", "testing-PHI-Gold-fixed")
```


```{r}
i2b2_input_filenames <- fs::dir_ls(i2b2_input_dir)
i2b2_gold_filenames <- fs::dir_ls(i2b2_gold_dir)

# Number of clinical notes in the evaluation dataset
length(i2b2_input_filenames)
length(i2b2_gold_filenames)
```
